---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# semFromKeys

<!-- badges: start -->
[![R-CMD-check](https://github.com/timbainbridge/semFromKeys/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/timbainbridge/semFromKeys/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The 'semFromKeys' package was designed to streamline running 'lavaan' models with similar structures using keys lists to generate model code instead of writing out the code for models manually. For confirmatory factor analyses (CFAs) and bi-factor models, the code creates and runs a series of models based on keys indicating each of the factors in the models. For exploratory factor analyses (EFAs) keys list are used to create a target rotation for a single EFA. For latent variable correlations, the model takes fitted CFA models and runs a series of models computing correlations between latent variables and, optionally, single items. For exploratory structural equation models (ESEM), there are two options. In each case, EFA factors predict scale factors. The first option (`esem.from.keys`) takes EFA and CFA keys as inputs and uses Rosseel and Loh's (2022) SAM method to prevent interpretational confounding (Burt, 1976), and the second (`esem.from.mods`) takes a fitted EFA model and fitted CFA and/or bi-factor models and uses Burt's (1976) 2-stage procedure to prevent interpretational confounding. The ESEM models were designed to run analyses analogous to those of Bainbridge, Ludeke, and Smillie (2022). Additionally, the `sem.path` function runs traditional latent variable structural equation models (SEM) using fitted CFA objects and path code as input.

Although the package might be of most use to those running ESEM similar to those of Bainbridge and colleagues (2022), it could also be very helpful to anyone wanting to create a correlation matrix based on latent variables rather than sum scores or to estimate a CFA measurement model for each scale in a sample to either check measurement characteristics before proceeding with further analyses or to simply compute measurement model based reliability statistics. It may also be useful to those wanting to preclude interpretational confounding in a standard latent variable SEM.

For sets of models that take a long time to run, code has been included to allow the first run to save outputs that can be checked against in subsequent runs. If nothing has changed, then the previous outputs are returned, saving the time (and energy) of running them again. To get this feature to work, the R version has to be 4.0 or later and a cache directory will have to be set with the `cache.setup()` function, which, by default, configures a cache directory in the users' cache as determined by the operating system. It can alternatively be set as a subdirectory within the current project or, if not using a project, the current working directory. Once the cache is set, `save_out = TRUE` can be included in function calls to save the relevant outputs, and `check = TRUE` can be included to look for previous outputs and only run models where something has changed. The feature means that small changes in data cleaning or model code need not necessitate re-running time-consuming models if only a small number have changed.

Given that the package enables creating files in a cache directory, the `cache.clean()` function has also been included to help clean up files. To comply with CRAN policies, the cache directory is set as a temporary environment variable, so it has to be set each time the global environment is cleared.

## Installation

You can install the development version of 'semFromKeys' from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("timbainbridge/semFromKeys")
```

You can install the stable CRAN version with:

``` r
install.packages("semFromKeys")
```

## Example

The following example generates keys, runs CFAs and an EFA using these keys, computes correlations between CFA latent variables, and uses outputs from these to run ESEMs. The alternative `esem.from.keys` function is also demonstrated.  

### CFAs

In this case, keys can be created from names in the dataset but they can also be created with simple code to generate a list.

```{r cfa-keys}
library(semFromKeys)
keys0 <- c("grit_c", "grit_p", "hope_a", "hope_p")
keys <- sapply(
  keys0, function(x) names(BFIGritHope)[grep(x, names(BFIGritHope))]
)
```

The lists should look something like this:

```{r cfa-keys2}
keys
```

Once keys are created, the CFAs can be run. The function produces messages of progress. These can help identify which models produced errors or warnings or to keep track of progress for collections of models with long run times.

```{r cfa-fit}
cfa_fit <- cfa.from.keys(keys, BFIGritHope, fit_save = TRUE)
```

Results can be examined; for example, standard 'lavaan' summaries: 

```{r example-cfa2}
lavaan::summary(cfa_fit$fit$grit_c)
```

And selected fit measures:

```{r example-cfa3}
cfa_fit$fit_measures[, c("cfi", "rmsea")]
```

These models can be used to examine the measurement characteristics of the scales or to calculate latent variable model-based reliability scores (e.g., with `sapply(cfa_fit$fit, function(x) semTools::compRelSEM(x)[[1]])` for composite reliability, Jöreskog, 1971).

Bifactor models can be run with a similar, albeit more complex, method. See `?bifactor.from.keys` for details.

### EFAs

As for CFAs, an EFA can be run from a keys list. In this case, the keys list indicates factor that items are expected to load on rather than separate models. These are used to generate a target rotation to help ensure the EFA matches expectations.

```{r ex}
keys_e0 <- paste0("bfi_", c("e", "a", "c", "n", "o"))
keys_e <- sapply(
  keys_e0,
  function(x) names(BFIGritHope)[grep(x, names(BFIGritHope))],
  simplify = FALSE
)
```

After the keys list has been created, the model can be run similarly to the CFAs. When running the model, fit measures can be restricted to speed up estimation if not all are required (as for 'lavaan's' `lavaan::fitMeasures()` function).

```{r example-efa}
efa_fit <- efa.from.keys(
  keys_e, BFIGritHope, check = FALSE, fit_save = TRUE,
  fit_measures = c("chisq", "df", "pvalue", "bic")
)
```

EFA results can be examined in a similar way to the CFAs.

```{r example-efa2}
# Not run due to length
# lavaan::summary(efa_fit$fit$efa)  # Standard lavaan summary
efa_fit$fit_measures                # Fit measures
```

### Correlations

The fitted CFA models can also be used to calculate latent variable correlations. In this example, Burt's 2-stage procedure is used by setting `nagy = FALSE` to save time, but in many cases Nagy and colleagues' (2017) method will be superior and is the default. See `?sem.cor()` for further details.

```{r example-lavaco}
latent_cors <- sem.cor(BFIGritHope, cfa_fit$fit, nagy = FALSE)
```

```{r example-lavaco-matrix}
round(latent_cors$cor_mat, 3)
```

### ESEM

Two function run ESEM with EFA factors predicting measurement model latent variables. The first of these---`esem.from.keys`---uses keys lists to run the models using Rosseel and Loh's (2022) "Structure After Measurement" (SAM) method, and the second---`esem.from.mods`---uses fitted CFA, bi-factor, and EFA models to run the models using Burt's (1976) 2-stage procedure. In general, the SAM method is superior, so `esem.from.keys` should be used whenever possible. However, `esem.from.keys` currently only supports single-factor measurement models and `esem.from.mods` should be used for bi-factor models. See `?esem.from.keys` for details.

```{r example-esam}
esam_fit <- esem.from.keys(BFIGritHope, keys_e, keys, fit_save = FALSE)
```

The function provides standard 'lavaan' outputs, as well as r-squared values and regression parameters.

```{r example-esam2}
lavaan::summary(esam_fit$fit$grit_c)  # lavaan summary for the SAM method
```

```{r example-esam3}
round(esam_fit$r2, 3)
```

```{r example-esam4}
esam_fit$b$grit_c
```

The `esem.from.mods` function provides the same outputs as `esem.from.keys`, only with somewhat biased standard error estimates.

```{r example-esem}
esem_fit <- esem.from.mods(
  BFIGritHope, efa_fit$fit$efa, cfa_fit$fit, fit_save = FALSE
)
```

```{r example-esem2}
# Not run due to length
# lavaan::summary(esem_fit$fit$grit_c)  # Standard lavaan summary
round(esem_fit$r2, 3)
```

```{r example-esem3}
esem_fit$b$grit_c
```

### SEM

Finally, `semFromkeys` includes a function to help run standard structural equation models (SEM). The function takes fitted CFA models for the measurement models and lavaan code for the structural path(s) and uses Rosseel and Loh's (2022) method to run the model.

```{r example-sem}
path <- "grit_p ~ grit_c + hope_p\nhope_p ~ hope_a"
sem_fit <- sem.path(path, BFIGritHope, cfa_fit$fit, fit_save = TRUE)
```

The function outputs include a standard lavaan summary, r-squared values, regression path coefficients and correlations, and fit measures.

```{r example-sem2}
print(summary(sem_fit$fit))      # Standard lavaan summary
```

```{r example-sem3}
sem_fit$b      # Standardised regression path coefficients
```

```{r example-sem4}
sem_fit$r2        # R^2 values
```

```{r example-sem5}
sem_fit$cors          # Correlations
```

```{r example-sem6}
# Fit measures
round(sem_fit$fit_measures[c("chisq", "df", "pvalue", "cfi", "rmsea")], 3)
```

## References

Bainbridge, T. F., Ludeke, S. G., & Smillie, L. D. (2022). Evaluating the Big Five as an organizing framework for commonly used psychological trait scales. Journal of Personality and Social Psychology, 122(4), 749-777. https://doi.org/10.1037/pspp0000395.

Burt, R. S. (1976). Interpretational confounding of unobserved variables in Structural Equation Models. Sociological Methods & Research, 5(1), 3-52. https://doi.org/10.1177/004912417600500101.

Jöreskog, K. G. (1971). Statistical Analysis of Sets of Congeneric Tests. Psychometrika, 36(2), 109-133. https://doi.org/10.1007/BF02291393.

Nagy, G., Brunner, M., Lüdtke, O., and Greiff, S. (2017). Extension Procedures for Confirmatory Factor Analysis. Journal of Experimental Education, 85(4), 574-596. https://doi.org/10.1080/00220973.2016.1260524.

Rosseel, Y. & Loh, W. W. (2022). A structural after measurement approach to structural equation modeling. Psychological Methods, 29(3), 561-588. https://doi.org/10.1037/met0000503.

<!-- You'll need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. -->
