---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# semFromKeys

<!-- badges: start -->
[![R-CMD-check](https://github.com/timbainbridge/semFromKeys/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/timbainbridge/semFromKeys/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The 'semFromKeys' package was designed to streamline running 'lavaan' models with similar structures using keys lists to generate model code instead of writing out the code for models manually. Currently, the code can run confirmatory factor analyses (CFAs), bifactor models, exploratory factor analyses (EFAs), latent variable correlations, and exploratory structural equation models (ESEMs). For CFAs and bifactor models, the code creates and runs a series of models based on keys indicating each of the factors in the models. For EFAs, keys list are used to create a target rotation for a single EFA. For latent variable correlations, the code takes fitted CFA models and runs models with correlations between all combinations or latent variables or selected latent variables with other selected latent variables. For ESEMs, the code takes a fitted EFA model and fitted CFA and/or bifactor models and runs an ESEM for each CFA or bifactor model input where the EFA factors predict the series of CFA/bifactor factors.

In the structural models (i.e., the latent variable correlations and the ESEM), Burt's (1976) 2-stage procedure is used to prevent interpretational confounding. The ESEM models were designed to run analyses equivalent to that of Bainbridge, Ludeke, and Smillie (2022).

For more sets of models that take a long time to run, code has been included to allow the first run to save outputs that can be checked against in subsequent runs. If nothing has changed, then the previous outputs are returned, saving the time (and energy) of running them again. To get this feature to work, the R version has to be 4.0 or later and a cache directory will have to be set with the `cache.setup()` function, which, by default, configures a cache directory in the users' cache as determined by the operating system. It can alternatively be set as a subdirectory within the current project or, if not using a project, the current working directory. Once the cache is set, `save_out = TRUE` can be included in function calls to save the relevant outputs, and `check = TRUE` can be included to look for previous outputs and only run models where something has changed.

Given that the package enables creating files in a cache directory, the `cache.clean()` function has also been included to help clean up files. To comply with CRAN policies, the cache directory is set as a temporary environment variable, so it has to be set each time the global environment is cleared.

## Installation

You can install the development version of 'semFromKeys' from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("timbainbridge/semFromKeys")
```

You can install the stable CRAN version with:

``` r
install.packages("semFromKeys")
```

## Example

The following example generates keys, runs CFAs and an EFA using these keys, and uses outputs from these to run ESEMs.

### CFAs

In this case, keys can be created from names in the dataset but they can also be created with simple code to generate a list.

```{r cfa-keys}
library(semFromKeys)
keys0 <- c("grit_c", "grit_p", "hope_a", "hope_p")
keys <- sapply(
  keys0, function(x) names(BFIGritHope)[grep(x, names(BFIGritHope))]
)
```

The lists should look something like this:

```{r cfa-keys2}
keys
```

Once keys are created, the CFAs can be run. The function produces messages of progress. These can help identify which models produced errors or warnings or to keep track of progress for collections of models with long run times.

```{r cfa-fit}
cfa_fit <- cfa.from.keys(keys, BFIGritHope, fit_save = TRUE)
```

Results can be examined. For example, standard 'lavaan' summaries: 

```{r example-cfa2}
lavaan::summary(cfa_fit$fit$grit_c)
```

And selected fit measures:

```{r example-cfa3}
cfa_fit$fit_measures[, c("cfi", "rmsea")]
```

These models can be used to examine the measurement characteristics of the scales or to calculate latent variable model-based reliability scores (e.g., with `sapply(cfa_fit$fit, function(x) semTools::compRelSEM(x)[[1]])` for composite reliability, Jöreskog, 1971).

### Latent variable correlations

It is also possible to examine correlations between the latent variables calculated above.

```{r example-lavaco}
latent_cors <- sem.cor(BFIGritHope, cfa_fit$fit)
latent_cors$cor_mat
```

### EFAs

As for CFAs, an EFA can be run from a keys list. In this case, the keys list indicates factor that items are expected to load on rather than separate models. These are used to generate a target rotation to help ensure the EFA matches expectations.

```{r ex}
keys_e0 <- paste0("bfi_", c("e", "a", "c", "n", "o"))
keys_e <- sapply(
  keys_e0,
  function(x) names(BFIGritHope)[grep(x, names(BFIGritHope))],
  simplify = FALSE
)
```

After the keys list has been created, the model can be run similarly to the CFAs. When running the model, fit measures can be restricted to speed up estimation if not all are required (as for 'lavaan's' `lavaan::fitMeasures()` function).

```{r example-efa}
efa_fit <- efa.from.keys(
  keys_e, BFIGritHope, check = FALSE, fit_save = TRUE,
  fit_measures = c("chisq", "df", "pvalue", "bic")
)
```

EFA results can be examined in a similar way to the CFAs.

```{r example-efa2}
# Not run due to length
# lavaan::summary(efa_fit$fit$efa)  # Standard lavaan summary
efa_fit$fit_measures                # Fit measures
```

### ESEM

Finally, outputs from these models can be used as inputs into ESEMs where the scales of the CFAs are regressed on the EFA factors.

```{r example-esem}
esem_fit <- esem.from.mods(
  efa_fit$fit$efa, cfa_fit$fit, data = BFIGritHope, fit_save = FALSE
)
```

The function provides standard 'lavaan' outputs, as well as r-squared values and regression parameters.

```{r example-esem2}
# Not run due to length
# lavaan::summary(esem_fit$fit$grit_c)  # Standard lavaan summary
round(esem_fit$r2, 3)
```

```{r example-esem3}
esem_fit$b$grit_c
```

To take advantage of functions' time-saving `check = TRUE` for subsequent running of code, a cache directory will need to be set. To see how to do this, see `?cache.setup`.

<!-- You'll need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. -->

## References

Bainbridge, T. F., Ludeke, S. G., & Smillie, L. D. (2022). Evaluating the Big Five as an organizing framework for commonly used psychological trait scales. Journal of Personality and Social Psychology, 122(4), 749-777. https://doi.org/10.1037/pspp0000395.

Burt, R. S. (1976). Interpretational confounding of unobserved variables in Structural Equation Models. Sociological Methods & Research, 5(1), 3-52. https://doi.org/10.1177/004912417600500101.

Jöreskog, K. G. (1971). Statistical Analysis of Sets of Congeneric Tests. Psychometrika, 36(2), 109-133. https://doi.org/10.1007/BF02291393.
